\chapter{Introduction}

The gravity that we know here on Earth is intuitive. It is familiar. It is
expected. And it is safe (for the most part). But this universe that we live in
continues to offer so much more than we expected. Starting from observing the
motion of the planets and stars in the night sky to theorizing about black
holes, the most compact and extreme objects in the universe, gravity continues
to humble us.

\section{Newtonian Gravity}

In order to explain black holes, the overarching topic of this thesis, we must
first start with a more simplified version of gravity and work our way up.
Newtonian gravity states that the gravitational force between two masses is
proportional to the two masses and inversely proportional to the square of the
distance between the masses, or
\begin{equation}\label{eq:fma}
   \mathbf{F} = -\frac{Gm_1m_2}{r^2}\mathbf{\hat{r}} ,
\end{equation}
with $G$ being Newtons constant. However, we usually don't just have point
masses like $m_1$ and $m_2$. We often have a distribution of matter which we'll
represent as a matter density $\rho$. This matter distribution will generate a
gravitational potential $\Phi$ via Poisson's equation
\begin{equation}\label{eq:poisson}
   \nabla^2\Phi = 4\pi G\rho .
\end{equation}

To relate Eqs. \ref{eq:fma} and \ref{eq:poisson}, we compute the gradient of
$\Phi$ as a force,
\begin{equation}
    \mathbf{F} = -\nabla\Phi.
\end{equation}
Then, because of Newtons generic force equations
\begin{equation}
   \mathbf{F} = m\mathbf{a} = m\frac{d^2\mathbf{x}}{dt^2},
\end{equation}
we can get the equations of motion for any object in Newtonian gravity,
\begin{equation}
   m\frac{d^2x_i}{dt^2} = -\frac{\partial \Phi}{\partial x_i} .
\end{equation}

With these equations, one is able to compute quite a lot of gravitational
scenarios. This could be the motion of a planet around a star or moving a
satellite to a Lagrange point of two orbiting bodies. Combined with some ideal
fluid mechanics for computing $\rho$, they can also be used to model the
interiors of large gas planets or even stars. These basic starting equations are
able to unlock a large amount of gravitational physics on their own, making them
vital, even today, to astrophysical research.

However, Newtonian gravity does have its limitations. One being how simultaneity
of events is measured. In order to understand simultaneous events, we must
understand the coordinates that are used and how they transform in Newtonian
gravity. If we take a step back, one way to conceptually think about Newtonian
gravity is that we have a time coordinate $t$. And at each time $t$, we have a
snapshot of our 3D world which is represented by some spatial coordinates
$x,y,z$. Therefore, two events that happen simultaneously will be labeled with
the same value for the time coordinate $t$, no matter where we are in our
spatial coordinates. However, we have some freedom with our spatial coordinates.
We can freely transform them from cartesian $(x,y,z)$ to spherical
$(r,\theta,\phi)$ to cylindrical $(s,\phi,z)$ or to any other coordinates we can
think of without altering our time coordinate, preserving simultaneity.
Conversely, the only way to transform our time coordinate (and keep the same
physics) is a simple $t\rightarrow t+\delta t$. Therefore, time and space are
somewhat decoupled in Newtonian gravity.

Another limitation of Newtonian gravity is that it doesn't account for the
finite speed of light. Related to the first limitation, events can only be
considered simultaneous if they occur at the same time $t$. Consider a scenario
where one person stands at rest at a point $(x,y,z)$ and time $t$ and shines a
flashlight in the $+x$ direction. A second person is moving with respect to the
first person in the $+x$ direction with a velocity $v$. This second person has
their own coordinate system $(t',x',y',z')$ different from the first person at
rest and would find the relation between the two systems to be
\begin{align}
   t' &= t \\
   x' &= x - vt \\
   y' &= y \\
   z' &= z .
\end{align}
However, what happens if the second person is moving very close to the speed of
light? (i.e. $v \approx c$). One may be tempted to think of this scenario as
similar to a different scenario where the first person throwing a baseball while
the second rides in a car. It'd be perfectly reasonable for a car to be moving
faster than a baseball. However, this is fundamentally impossible with light.
The speed of light is constant in all coordinate systems. So if we try to
interpret this scenario in a Newtonian way, the second person would be moving
almost in line with the front of the light beam at the same speed. But then what
would happen if this second person shined their own flashlight? Would the this
second beam of light be moving almost twice as fast as the first one? Would it
stay effectively stationary and not go anywhere because the second person was
already moving at the speed of light? This inconsistency and impossibility is
why Newtonian gravity isn't the full picture and why we need a different theory.

\section{Special Relativity}

The theory of special relativity (SR) is able to reconcile and account for the
finite speed of light in all coordinate systems. Say we have some event $P$.
Fig. \ref{fig:light_cone} shows a light cone centered at event $P$. The vertical
axis is time and the horizontal axis is a spatial dimension. The lines at
$\pm45^{\circ}$ represent the paths that light take both into the future and
from the past. In the ``Past'' light cone, everything contained within the light
cone is able to influence event $P$ in a causal way (i.e. things traveling at or
slower than the speed of light). Everything within the ``Future'' light cone can
be influenced by event $P$. Events within these light cones are said to be
causally connected to $P$. However, everything outside the light cone is unable
to affect or be influence by event $P$, thus we call those events causally
disconnected.

\begin{figure}[h]
   \centering
   \begin{tikzpicture}
      % Future cone fill — bounded by straight edges and upper arc of top ellipse
      \fill[gray!20] (0,0) -- (2,2)
         arc[start angle=0, end angle=180, x radius=2, y radius=0.5]
         -- cycle;
      % Past cone fill — bounded by straight edges and lower arc of bottom ellipse
      \fill[gray!20] (0,0) -- (2,-2)
         arc[start angle=0, end angle=-180, x radius=2, y radius=0.5]
         -- cycle;
      % Cone straight edges
      \draw[thick] (0,0) -- (2,2);
      \draw[thick] (0,0) -- (-2,2);
      \draw[thick] (0,0) -- (2,-2);
      \draw[thick] (0,0) -- (-2,-2);
      % Top ellipse: solid upper arc (visible rim), dashed lower arc (hidden rim)
      \draw[thick] (2,2)
         arc[start angle=0, end angle=180, x radius=2, y radius=0.4];
      \draw[thick] (2,2)
         arc[start angle=0, end angle=-180, x radius=2, y radius=0.25];
      % Bottom: only the front arc (curved base of past cone, not open)
      \draw[thick] (2,-2)
         arc[start angle=0, end angle=-180, x radius=2, y radius=0.5];
      % Center point
      \filldraw[black] (0,0) circle (3pt) node[right=4pt] {$P$};
      % Labels
      \node at (0, 2.9) {Future};
      \node at (0,-2.9) {Past};
      % Axes (left side)
      \draw[-{Triangle[length=5pt,width=4pt]}, thick] (-3.2,-0.6) -- (-3.2,0.6) node[above] {$t$};
      \draw[-{Triangle[length=5pt,width=4pt]}, thick] (-3.2,-0.6) -- (-2.0,-0.6) node[right] {$x$};
   \end{tikzpicture}
   \caption{A light cone centered at the event $P$. The cone edges represent
   rays propagating at the speed of light $c$.}
   \label{fig:light_cone}
\end{figure}

In Newtonian gravity, we have a simple notion of an invariant distance between
two points, e.g.
\begin{equation}
   (\Delta s)^2 = (\Delta x)^2 + (\Delta y)^2 + (\Delta z)^2.
\end{equation}
This measure of distance is independent of what coordinate system we use for our
spatial coordinates and works for Newtonian gravity because all time coordinates
of all observers are the same. There is a universal time in Newtonian gravity.
However, in SR we no longer have this universal time which everybody can agree
on. Therefore, the measure of ``distance'' that we use in SR must account for
our time coordinate somehow because of the coupling between space and time that
the finite speed of light demands. For now, we just present this distance
measure, called the spacetime interval, as
\begin{equation}\label{eq:spacetime-interval}
   (\Delta s)^2 = -(\Delta t)^2 + (\Delta x)^2 + (\Delta y)^2 + (\Delta z)^2.
\end{equation}

First, notice that the units of this don't match. We are adding something like
(seconds)$^2$ to something like (meters)$^2$. But one can notice that if we
multiply the $\Delta t$ by the speed of light $c$, we'll get the proper units.
Very quickly in relativity, it becomes annoying to always have to write and keep
track of this factor of $c$. Therefore, almost everybody in the field simply
sets $c=1$ with the assumption that if we ever need to calculate actual
quantities, we'll add it back in at the end.

Second, we will now use this spacetime interval to motivate constructing
something called a spacetime ``metric''. Now to fully understand the significance of what
a metric is and what it represents requires taking courses on differential
geometry. However, for our purposes, a metric is basically a $4\times 4$ matrix
that represents how coordinates combine together to form the spacetime interval.
Therefore, for the interval in Eq. \ref{eq:spacetime-interval}, we can define
its spacetime metric, named the Minkowski metric, as
\begin{equation}
   \eta_{\mu\nu} =\left(
   \begin{matrix}
      -1 & 0 & 0 & 0 \\
       0 & 1 & 0 & 0 \\
       0 & 0 & 1 & 0 \\
       0 & 0 & 0 & 1
   \end{matrix}
   \right).
\end{equation}
It is diagonal because there are no cross-terms in Eq.
\ref{eq:spacetime-interval} (such as $\Delta x\Delta y$). We can then rewrite
Eq. \ref{eq:spacetime-interval} using Einstein summation notation (where all
repeated indices are implicitly summed)
\begin{equation}
   (\Delta s)^2 = \eta_{\mu\nu}\Delta x^\mu\Delta x^\nu
\end{equation}
where
\begin{equation}
   x^\mu =\left(
   \begin{matrix}
      t \\
      x \\
      y \\
      z
   \end{matrix}
   \right).
\end{equation}

With a spacetime metric, we are able to do so many things. We can raise and
lower indices of different quantities (i.e. transform between vectors and dual
vectors/one-forms, $x_\nu = \eta_{\mu\nu}x^\mu$). We can turn the spacetime
interval from $\Delta$s into infinitesimals and integrate it to find the length
of a spacetime curve. But one of the most powerful things we can do is define
transformations between coordinates that leave spacetime intervals unchanged
(similar to Newtonian gravity).
Lets define a transformation from an un-primed coordinate system to a primed one
with
\begin{equation}
   x^{\mu'} = {\Lambda^{\mu'}}_{\mu} x^\mu .
\end{equation}
To show that this leaves the spacetime interval unchanged, let's plug it in
\begin{align}
   (\Delta s)^2 &= \eta_{\mu\nu}\Delta x^\mu\Delta x^\nu \\
   &= \eta_{{\mu'}{\nu'}}\Delta x^{\mu'}\Delta x^{\nu'} \\
   &= \eta_{{\mu'}{\nu'}}{\Lambda^{\mu'}}_{\mu}\Delta x^{\mu}{\Lambda^{\nu'}}_{\nu}\Delta x^{\nu} \\
   &= \eta_{{\mu'}{\nu'}}{\Lambda^{\mu'}}_{\mu}{\Lambda^{\nu'}}_{\nu}\Delta x^{\mu}\Delta x^{\nu}.
\end{align}
Therefore, $\Lambda$ is a transformation between coordinates only if
\begin{equation}
   \eta_{\mu\nu} = \eta_{{\mu'}{\nu'}}{\Lambda^{\mu'}}_{\mu}{\Lambda^{\nu'}}_{\nu},
\end{equation}
Or said another way, $\Lambda$ doesn't transform the components of the metric.
If $\Lambda$ satisfies this, then we call it a Lorentz transformation. These
will be mentioned in Ch. \ref{ch:compare}.

\section{General Relativity}
Now in going from Newtonian gravity to SR, we didn't actually add in any gravity
to SR. We only tackled the issues of simultaneity and the speed of light. Now we
are ready to combine SR with gravity to produce Einstein's theory of general
relativity (GR). Probably the biggest conceptual difference between GR and
Newtonian gravity is that Newtonian gravity is explained as a force exerted
between two masses. And any object that has a force being applied to it will
accelerate in the direction of that force. Therefore, motion in Newtonian
gravity arises from forces.

In GR, however, motion does not arise from forces. Instead, in GR we say there
is some ``background'' spacetime, or some fabric of the universe that matter
exists both on and in at the same time. In the Newtonian and SR picture, this
background is ``flat'' meaning that it doesn't affect the matter that lives in
it. However, in GR this background doesn't have to be flat anymore. It can be
distorted, or curved; sort of how a trampoline with a heavy object in the middle
is curved. Motion then arises not because of any force, but simply due to the
curvature of the background spacetime. The motion of a small mass is dictated by
how much spacetime is curved. In the trampoline example, a small mass starting
with a constant velocity on a flat spacetime (flat trampoline) will not deviate
from its original path and will continue with a constant velocity. That same
mass moving on a curved spacetime (trampoline with a heavy object in the middle)
will naturally move towards the heavy object with increasing velocity without
any outside influence. It's the nature of the curved spacetime which drives the
motion of matter. And the beauty of Einstein's theory is that it is matter
itself which drives the curvature of spacetime.

To describe this curvature, similar to SR, GR will also have a metric associated
with it. We will denote this metric as $g_{\mu\nu}$ because $\eta_{\mu\nu}$ is
reserved for Minkowski spacetime. Our new metric $g_{\mu\nu}$ will encode
everything we need to know about the gravity of our spacetime.

We first start by looking back at Poisson's equation for a gravitational
potential (rewritten here for ease)
\begin{equation}
   \nabla^2\Phi = 4\pi G\rho .
\end{equation}

\begin{enumerate}
   \item Broad intro to gravity
   \begin{enumerate}
      \item Newtonian Gravity
      \begin{enumerate}
         \item poisson equation
         \item equations of motion (aka f=-gradPhi=ma)
         \item simultaneity
      \end{enumerate}
      \item SR
      \begin{enumerate}
         \item flat space interval
         \item metric
         \item lortentz transformations
      \end{enumerate}
      \item GR
      \begin{enumerate}
         \item how vectors move in a curved space
         \item covariant derivatives,parallel transport
         \item metric
         \item connection
         \item geodesics
         \item compare with newton
         \item Riemann
         \item Ricii
         \item Einstein tensor
         \item Conservation of energy
         \item Stress energy tensor
         \item relate poisson equation to Einstein field equations
         \item vacuum Einstein's equations
      \end{enumerate}
      \item NR (GR on computers, different formulations)
      \begin{enumerate}
         \item Need to foliate spacetime
         \item Coordinates
         \item spatial metric, lapse, and shift
         \item 3+1 decomposition
         \item Gauge freedoms
         \item BSSN
         \item GH
      \end{enumerate}
   \end{enumerate}
   \item Ultimate goals of Numerical Relativity (models/training)
   \item Why HPC to achieve goals
   \item GH and why horizon finding is important
   \item Chapters of thesis
   \begin{enumerate}
      \item Brief on Lovelace and Nelli et al.
      \item Brief on Nelli et al.
      \item Brief on Cross code comparison
   \end{enumerate}
\end{enumerate}
