\chapter{Infrastructure for cross-code comparison of CCE waveforms\label{ch:compare}}

\newcommand{\todocite}{\todo{CITE}}

% Use bibunit if using bibtex/natbib
\begin{bibunit}[unsrtnat]

\section{Introduction}

Being able to compare gravitational waveforms with each other, whether it be
comparing Numerical Relativity (NR) waveforms with other NR waveforms or
observed gravitational waves (GWs), is an extremely important ability in GW
science; potentially the most important. GW detectors compare suspected signals
to models that have been informed by NR such as the IMRPhenom family
(examples: \cite{Pratten:2020fqn, Garcia-Quiros:2020qpx, Pratten:2020ceb}) for
frequency-domain analysis and the SEOBNR family (examples:
\cite{Khalil:2023kep, Ramos-Buades:2023ehm, vandeMeent:2023ols}) for time-domain
analysis. NR codes also compare waveforms internally, typically for verification
of convergence as in \S\ref{sec:gws} or for analysis of trends in GW catalogs
\cite{Scheel:2025jct}. However, comparing waveforms from different NR
simulations, even from the same code, have to be done carefully since different
simulations will be done in different gauges.

\subsection{Electricity and Magnetism example}

To give an example of why we need to consider gauges when comparing NR
waveforms, let us first look at a more familiar use of gauges in an
introductory setting such as electricity and magnetism. Maxwell's equations (in
vector form) are
\begin{align}
  \nabla \cdot \mathbf{E} = 4\pi\rho &\quad\quad
  \nabla \times \mathbf{E} = -\frac{1}{c}\frac{\partial \mathbf{B}}{\partial t} \\
  \nabla \cdot \mathbf{B} = 0 &\quad\quad
  \nabla \times \mathbf{B} = \frac{1}{c}\left(4\pi\mathbf{J} + \frac{\partial \mathbf{E}}{\partial t}\right)
\end{align}
From Gauss's law for magnetism and Faraday's equation along with vector calculus
identities, we can define
\begin{align}
  \mathbf{B} &= \nabla \times \mathbf{A}\label{eq:maxwell-b} \\
  \mathbf{E} &= -\nabla\phi - \frac{\partial \mathbf{A}}{\partial t}\label{eq:maxwell-e}
\end{align}
where $\phi$ is the electric potential and $\mathbf{A}$ is the vector potential.
We notice that if we modify $\phi$ and $\mathbf{A}$ like so
\begin{align}
  \mathbf{A} &\rightarrow \mathbf{A} + \nabla\psi\label{eq:maxwell-a} \\
  \phi &\rightarrow \phi - \frac{\partial \psi}{\partial t}\label{eq:maxwell-phi}
\end{align}
for some scalar function $\phi$, $\mathbf{B}$ and $\mathbf{E}$ in
\ref{eq:maxwell-b} and \ref{eq:maxwell-e}, respectively, remain unchanged. This
ability to choose $\phi$ and $\mathbf{A}$ is called gauge freedom, and the
choice that we make is considered the gauge.

In electricity and magnetism there are two common gauges; Coulomb and Lorenz.
The Coulomb gauge is defined by
\begin{equation}
  \nabla \cdot \mathbf{A} = 0
\end{equation}
To describe the Lorenz gauge, we combine $\phi$ and $\mathbf{A}$ as a
four-vector
\begin{equation}
  A^\mu = \left(\phi/c,\mathbf{A}\right) .
\end{equation}
The Lorenz gauge is then
\begin{equation}
  \partial_\mu A^\mu = 0.
\end{equation}
This differs from the Coulomb gauge because of the $\partial_0 A^0 \sim
\partial_t \phi$ part. The implications that this choice of gauge has on the
equations can be found in virtually any electricity and magnetism textbook
\todocite. The important point for this discussion is that even though
$\mathbf{E}$ and $\mathbf{B}$ mathematically remain unchanged depending on which
gauge is chosen, the gauge dependent quantities $\phi$ and $\mathbf{A}$ cannot
be compared between gauges.

\subsection{General and Numerical Relativity gauges}

Now we move to gauges in general relativity (GR). Instead of putting a
constraint on the four-potential as our gauge condition, in GR we impose our
gauge condition on the coordinates $x^\mu$ of our spacetime. We skip a
derivation of \textit{why} we have this coordinate freedom in the theory of
general relativity as that's the topic of introductory GR courses, but a
derivation can be found in virtually any GR text book, including
\cite{Misner:1973prb,Wald:1984rg,Carroll:2004st} to name a few. For our
purposes, let's take a look at a few gauges in GR and NR.

\subsubsection{Transverse-traceless gauge}
The first gauge most students encounter is one called the transverse-traceless
(TT) gauge, occasionally called the radiation gauge. The TT gauge is
particularly useful in a linearized theory of gravity where we define our metric
as
\begin{equation}\label{eq:linear_gravity}
  g_{\mu\nu} = \eta_{\mu\nu} + \bar{h}_{\mu\nu}
\end{equation}
with
\begin{equation}
  \bar{h}_{\mu\nu} = h_{\mu\nu} - \frac{1}{2}\eta_{\mu\nu}h
\end{equation}
where $\eta_{\mu\nu}$ is our flat background spacetime metric and $h_{\mu\nu}$
is a small perturbation of this background. Here, $\bar{h}_{\mu\nu}$ is a
``trace-reversed'' form. We take our field equations
\begin{equation}\label{eq:field_equations}
  R_{\mu\nu} - \frac{1}{2}Rg_{\mu\nu} = 8\pi T_{\mu\nu} 
\end{equation}
and substitute Eq. \ref{eq:linear_gravity} into them. This will contain second
derivatives of $\bar{h}_{\mu\nu}$ because of the Ricci tensor and scalar. It is at
this point where we make a gauge choice of
\begin{equation}\label{eq:harmonic_gauge}
  \Box x^\mu = 0
\end{equation}
which can be shown to simplify to \cite{Carroll:2004st}
\begin{equation}\label{eq:linear_harmonic_gauge}
  \Box x^\mu = 0 \rightarrow g^{\mu\nu}\Gamma^\rho_{\mu\nu} = 0 \rightarrow
  \partial_\mu {\bar{h}^\mu}_{\lambda} = 0
\end{equation}
in the weak-field (linear) limit. Plugging Eq. \ref{eq:linear_harmonic_gauge}
into Eq. \ref{eq:field_equations} and taking the vacuum case ($R_{\mu\nu} =
T_{\mu\nu} = 0$), we arrive at a clean result
\begin{equation}
  \Box \bar{h}_{\mu\nu} = 0 .
\end{equation}
It can be shown (through some math that is unimportant to the point here)
\cite{Carroll:2004st} that
\begin{equation}
  h^{\textrm{TT}}_{\mu\nu} = \left(
\begin{matrix}
  0 & 0 & 0 & 0 \\
  0 & h_{+} & h_{\times} & 0 \\
  0 & h_{\times} & -h_{+} & 0 \\
  0 & 0 & 0 & 0
\end{matrix}
\right) ,
\end{equation}
where we have dropped the bars because $h_{\mu\nu}$ is the trace-reverse of
$\bar{h}_{\mu\nu}$ which is traceless. These are the familiar GW polarizations
$h_{+}$ and $h_{\times}$ for a linear GW propagating in the z-direction.
Therefore, this is a useful gauge to analyze GWs from one of the detectors
(though it's quite a bit more complicated than this in reality
\cite{KAGRA:2021vkt,LIGOScientific:2021usb,LIGOScientific:2020ibl,LIGOScientific:2018mvr}).

\subsubsection{NR gauges}

Moving on to NR, the two most popular formulations of the field equations (Eq.
\ref{eq:field_equations}) are the Generalized Harmonic (GH) formulation
\cite{Lindblom:2005qh} and the BSSN/Z4 family \todocite. Each of these
formulations use different gauge conditions for the evolution of the coordinates
$x^\mu$, none that are equal to the TT gauge.

We start with GH. There are two dominant gauges used when evolving the GH
formulation. The first is actually the harmonic gauge introduced in Eq.
\ref{eq:harmonic_gauge}. The TT gauge is actually a sub-category of the harmonic
gauge, useful for linear theories. However, GH is the fully non-linear theory of
gravity and so the TT gauge is not suitable here. The harmonic gauge is
convenient because the solutions for the equations of the coordinates are
harmonic (or wave) solutions, hence the name, and we know how to solve for
waves.

The second gauge commonly used in GH evolutions is one that actually builds off
of the harmonic gauge, because it's wave solutions are so desirable. The general
form of this gauge, called the damped-harmonic gauge, is
\begin{equation}
  \Box x_\mu = H_\mu(x^\nu, g_{\nu\kappa})
\end{equation}
where $H_\mu$ is called the gauge-source function~\cite{Szilagyi:2009qz}, and it
is a function of the coordinates and the spacetime metric. The gauge source
function is defined in Eq. \ref{eq:dh_gauge_source}--\ref{eq:dh_mu3}. The main
purpose for this is to damp away oscillations shift $\alpha$ and drive it to one
(so the time evolution remains steady).

For the BSSN/Z4 family of formulations, a gauge condition called the
moving-puncture gauge, combined with $1+\log$ slicing and a
$\Gamma$-driver condition, is used. For this family, conditions are put onto the
lapse function $\alpha$ and the shift vector $\beta^i$ instead of having
equations for the coordinates themselves. The $1+\log$ slicing condition is (Ch.
4 of \cite{Baumgarte:2010ndz})
\begin{equation}
  \partial_t \alpha = -2\alpha K + \beta^j\partial_j\alpha
\end{equation}
where $K$ is the extrinsic curvature. The $\Gamma$-driver condition can come
in two forms (Ch. 4 of \cite{Baumgarte:2010ndz})
\begin{equation}
  \partial_t \beta^i = \eta B^i + \beta^j\partial_j\beta^i,
\end{equation}
or (\cite{Hilditch:2012fp})
\begin{equation}
  \partial_t \beta^i = \tilde{\Gamma}^i - \eta\beta^i + \beta^j\partial_j\beta^i,
\end{equation}
where $\eta$ is a tunable parameter, and $\tilde{\Gamma}^i$ is related to the
Christoffel symbols. The former is typically used with BSSN while the later is
used with the Z4 family, but many NR codes make slight modifications to any of
these gauge equations.

It should be fairly clear that trying to compare a GW produced with the GH
formulation to a GW produced with the BSSN formulation is not well-defined
because the gauges that are used are completely different. However, more subtly,
comparing GWs that were produced with the same formulation is also ill-defined.
This is due to the limitations of numerical resolution and floating-point
arithmetic when solving Einstein's Equations on the computer. Because of these
numerical issues, even simulations that start with the same initial data but are
evolved with different spatial or temporal resolutions may not evolve in an
identical manner. Therefore, the coordinates of these different resolution
simulations will be different at the same ``time'' of the simulation. This can
be seen in the bottom two panels of Figure \ref{fig:BbhGWs} where we are
comparing the phase and amplitude difference of the GWs from different
resolutions of BBH simulations evolved with the same initial data.

\section{CCE?}

Now that we have detailed a number of different useful gauges, namely the TT,
harmonic, damped harmonic, and moving-puncture gauges, it should be clear that
the GW strains $h$ we compute from simulations done in these different gauges
cannot directly be compared with each other. The coordinate freedom of GR
implies that even if we use the same formulas to compute $h$, we will get
different answers because the coordinates are completely different. Therefore,
how can we compare waveforms across codes? The answer is, by comparing the
waveforms at a location where our choice of coordinates mostly don't matter;
Namely at future-null infinity, also denoted \scri. Future-null infinity is the
location in spacetime that future null rays (e.g. light rays) escape if they
don't terminate inside the black hole and aren't in a stable orbit.

There are a few ways of computing GWs at \scri that allow for comparison of GWs
between resolutions or between codes. Most NR codes don't have a computational
domain that extends out to \scri, as this would be intractably computationally
expensive with typical methods. Instead, GWs at \scri must be computed in
another manner. The first, and by far most inaccurate, method is to simply
ignore gauge effects and compare GWs that are extracted at finite radii within
the computational domain, treating them as if they were at \scri. While this
method cannot be used to compare waveforms between NR codes, when comparing
similar simulations from the same code, it can be a quick-and-dirty method for a
quick analysis.

The second, and by far most common, method for computing GWs at \scri is called
extrapolation. The details for extrapolation can be found in the SXS
Collaboration's second catalog release \cite{Boyle:2019kee}. However, the overall
method is to compute the strain $h$ and the $\Psi_4$ Weyl scalar on a number of
concentric spheres surrounding the origin of the coordinate system. These
quantities are then extrapolated with a $1/r^N$ power law, where $N$ is chosen
based on a number of different conditions described in \cite{Boyle:2019kee}.

Extrapolation has been very successful and has been widely used by the NR
community for computing GWs \todocite. However, because it simply extrapolates
waveforms from close to the black hole, some non-linear effects of GR as the
wave travels to \scri are not accounted for and may be important. Even with
memory corrected waveforms from \cite{Scheel:2025jct}, they still don't account
for all non-linear effects. For this, a full evolution of Einstein's equations
out to \scri is needed.

\subsection{Also CCE?}

A method to evolve Einstein's equations all the way out to \scri is called
Cauchy-Characteristic Extraction/Evolution (CCE). Here, we detail the important
concepts and equations of CCE that will be needed in the rest of the chapter,
following \cite{Moxon:2020gha} and \cite{Moxon:2021gbv}.

CCE works by first performing a Cauchy evolution of the spacetime using one of
the formulations we discussed earlier this chapter (GH, BSSN, Z4). Then, we
store the spacetime metric and its derivatives on a sphere centered at the
origin for all times during our simulation. This is called the worldtube. This
worldtube data is then used as the inner boundary condition for the
Characteristic evolution which goes out to \scri. This characteristic evolution
is performed on null-hypersurfaces, unlike the Cauchy evolution which is
performed on space-like hypersurfaces. This is because with the Characteristic
evolution, we are primarily concerned with evolving and resolving GWs, which
travel along null rays, making null-hypersurfaces well suited to this problem.

CCE decomposes the metric into again different coordinates and gauge conditions
that GH or BSSN. The metric in a Bondi-like spherical coordinate system
$\{u,r,x^A\}$ is
\begin{equation}
  ds^2 = -\left(e^{2\beta}\frac{V}{r} - r^2h_{AB}U^AU^B\right)du^2 - 2e^{2\beta}dudr
    -2r^2h_{AB}U^Bdudx^A + r^2h_{AB}dx^Adx^B
\end{equation}
where $h_{AB}$ is the related to the metric on the 2-sphere $q_{AB}$, $V$ and
$\beta$ are scalar functions, and $U^A$ is a vector on the 2-sphere. These
Bondi-like coordinates have the gauge conditions that $g_{rr} = g_{rA} = 0$,
$\det{(h_{AB})} = \det{(q_{AB})}$, and asymptotic behavior
\begin{equation}
  \lim_{x\rightarrow\infty} \{\beta, U^A, h_{AB}, \frac{V-r}{r^2} \} = \mathcal{O}(r^0) .
\end{equation}
We can define a complex dyad $q^A$ (basis vector) of $q_{AB}$ as the following
\begin{equation}
  q_{AB} = \frac{1}{2}\left(q_A\bar{q}_B + \bar{q}_Aq_B\right),
\end{equation}
with barred quantities being the complex conjugate. Additionally, if a quantity
transforms like
\begin{equation}
  q_A \rightarrow q_A e^{is\psi}
\end{equation}
for some phase $\psi$, then we say that quantity has spin weight $s$. With these
definitions, it is now convenient to define some scalars that will be important
for the CCE evolution:
\begin{align}
  \label{eq:bondi_u}
  U &\equiv U^Aq_A & (s=1) \\
  Q &\equiv r^2e^{-2\beta}q^Ah_{AB}\partial_r U^B & (s=1) \\
  r^2W &\equiv V-r & (s=0) \\
  J &\equiv \frac{1}{2}q^aq^Bh_{AB} & (s=2) \\
  K &\equiv \sqrt{1-J\bar{J}} & (s=0) \\
  \label{eq:bondi_h}
  H &\equiv \partial_u J & (s=2)
\end{align}

Now we can define a hierarchical set of evolution equations for these scalars
which will give us the characteristic evolution of the spacetime on null-slices.
The first of these equations that is needed is actually Eq. \ref{eq:bondi_h},
which will give us $J$. Then once we have $J$, we can build up a solution by
solving radial differential equations for the other scalars. Here we will denote
the right hand side of each equation with a generic function $S$ because the
details of what is on the right hand side are not important. See
\cite{Moxon:2020gha} again for the full equations.
\begin{align}
   \partial_r\beta &= S_\beta(J), \\
   \partial_r (r^2Q) &= S_Q(J,\beta), \\
   \partial_r U &= S_U(J,\beta,Q), \\
   \partial_r (r^2W) &= S_W(J,\beta,Q,U), \\
   \partial_r (rH) + L_H(J,\beta,Q,U,W)H &\\
   + L_{\bar{H}}(J,\beta,Q,U,W)\bar{H} &= S_H(J,\beta,Q,U,W)
\end{align}
The final radial equation for $H$ lets us solve Eq. \ref{eq:bondi_h} for J at
the next time step and the hierarchy starts over again.

Now these scalars are defined in terms of our metric expressed in a Bondi-like
gauge which is useful for evolving the Characteristic system. However, in order
for us to impose our worldtube data which is in some arbitrary evolved gauge, we
must transform the evolved gauge into our Bondi-like gauge. The mathematics of
this transformation are quite tedious and unimportant for the point being made
here because it uses a number of intermediate coordinates that are only helpful
for the transformation, so we will skip them. To see them in detail, refer to
\cite{Moxon:2020gha}. Instead we just state that our scalars defined in
\ref{eq:bondi_u}-\ref{eq:bondi_h} can be expressed as some function of the
Cauchy spacetime metric and Cauchy evolved coordinates (e.g. $U =
f(x^\mu,g_{\mu\nu})$, where $x^\mu$ is not related to $x^A$), with all the
coordinate transformations hidden.

\subsection{Machinery, Figure out section name later}

Now we have the basics of the Characteristic evolution. Specifically that we use
the Bondi scalars defined in Eqs. \ref{eq:bondi_u}-\ref{eq:bondi_h} as part of a
hierarchical evolution system, we need those Bondi scalars on the worldtube
which serve as the inner boundary condition for the Characteristic evolution,
and these Bondi scalars on the worldtube can be computed from the Cauchy metric,
coordinates, and the spatial and temporal derivatives of the metric. Next we
focus on how this last step is implemented in \spectre~and what new features
were added in \spectre.

The first detail we need to discuss is how \spectre~actually represents and
stores the Bondi scalars either in memory of the simulation or on disk. Since we
need the Bondi scalars on a sphere of constant radius at each timestep,
representing the scalars as complex spherical harmonic coefficients is the most
natural. For a surface $S_{\ell,m}$, we require $m=\ell$. This implies we will
use $M = (\ell+1)^2$ complex-valued coefficients (or modes) to represent each
Bondi scalar. We will call this the ``modal'' representation. The modal
representation is extremely useful for two reasons. First is that it is straight
forward to interpolate between two representations with different $\ell$ since
the interpolation is either truncating the number of coefficients or appending
coefficients with value zero. Second, is that the modal representation is
significantly more compact that a ``nodal'' representation where we represent
the complex valued scalar at collocation points on a sphere. For the same
surface $S_{\ell,m}$ the number of nodes needed to represent the scalar is $N =
(\ell+1)(2\ell+1) > M$. The advantage with using a nodal representation for the
actual Characteristic evolution is that we can easily perform arithmetic between
Bondi scalars. This is significantly more difficult with a modal representation
because you cannot perform arithmetic with coefficients.

We now have a situation where it is ideal to represent the evolved Bondi
scalars with nodal values, but the worldtube Bondi variables with modal
coefficients. The worldtube variables must be written to disk during the Cauchy
evolution, so we want those to take up as little disk space as possible. Also,
there's no guarantee that the $\ell$ we write the worldtube variables with will
be the same $\ell$ that we use for the inner boundary of our Characteristic
evolution. One may even want to run multiple Characteristic evolutions with
varying resolution.

\begin{itemize}
  \item Modal vs nodal
  \begin{itemize}
    \item Modal allows for interpolation
    \item Modal less data on disk
    \item Nodal is used for evolution (because DG)
  \end{itemize}
  \item $g_{\mu\nu}$ in ADM variables with time derivs, for old spec
  \item Bondi scalars directly from spectre
  \item Other codes
  \begin{itemize}
    \item Nodal easier for output
    \item Some codes can do modal, less data on disk
    \item Uses ADM variables but cannot output first time derivs
    \item Bondi nodal for completeness?
  \end{itemize}
  \item \spectre~only accepts Bondi modal
  \begin{itemize}
    \item More straight forward for time interpolation
    \item Create conversion exe to Bondi modal
    \item Documentation
  \end{itemize}
  \item Offer pre-built exes for people to download
\end{itemize}

Particularly, we specify that the scalars \ref{eq:bondi_u}-\ref{eq:bondi_h} are
defined in terms of the ADM metric variables $\alpha$, $\beta^i$, $g_{ij}$, and
their derivatives (both temporal and spatial cartesian).


\begin{itemize}
  \item Need to get to ADM modal form. We have machinery for that
  \item Input that were already supported
  \begin{itemize}
    \item Bondi modal
    \item Metric modal (dumping all time derivatives because of GH)
  \end{itemize}
  \item Input that I added
  \begin{itemize}
    \item Bondi nodal
    \item Metric nodal
    \item Metric modal (with BSSN/Z4 time derivative compute)
  \end{itemize}
\end{itemize}


\begin{enumerate}
  \item Motivation
  \begin{enumerate}
      \item Explain importance about comparing waveforms
      \item Waveform models, surrogates, etc.
      \item $h_{+}$ and $h_{\times}$ not gauge invariant
      \item Analog to 4 potential from E\&M
      \item Different evolution gauges
  \end{enumerate}
  \item Overview of CCE (from Jordan's paper)
  \item Different formulations evolved variable
  \item Converting to ``standard'' variables for CCE
  \item Caveat of frame fixing/BMS transforms
\end{enumerate}

\section*{References}
\putbib[cross_code_cce]

\end{bibunit}